---
title: "miaSim: Microbiome Data Simulation"
date: "`r Sys.Date()`"
package: 
    miaSim
output: 
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: yes
        toc_depth: 2
        number_sections: true
vignette: >
    %\VignetteIndexEntry{miaSim}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    \usepackage[utf8]{inputenc}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(cache = FALSE,
                        fig.width = 9,
                        message = FALSE,
                        warning = FALSE)
```

`miaSim` implements tools for microbiome data simulation based on the
`SummarizedExperiment` [@SE], `microsim` .     

# Installation

### Bioc-release

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

    BiocManager::install("miaSim")
```

### Bioc-devel

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# The following initializes usage of Bioc devel
BiocManager::install(version='devel')

BiocManager::install("miaSim")
```

# Models for simulating microbiome data sets 

The generalized Lotka-Volterra simulation model fitted to time-series estimates
microbial population dynamics and relative rates of interaction.

The Hubbell Neutral simulation model explains the diversity and relative 
abundance of species in ecological communities.

The Self-Organised Instability (SOI) model generates time series for communities
and accelerates stochastic simulation.

The simulations result in abundance matrix which can be used to form 
SummarizedExperiment object.

# Session info

```{r}
sessionInfo()
```
