---
title: "miaSim: Microbiome Data Simulation"
author:
- name: Daniel Rios Garza
  email: danielrios.garza@kuleuven.be
- name: Emma Gheysen
  email: emma.gheysen@student.kuleuven.be
- name: Karoline Faust
  email: karoline.faust@kuleuven.be
- name: Leo Lahti
  email: leo.lahti@iki.fi
- name: Yagmur Simsek
  email: yagmur.simsek@hsrw.org
- name: Yu Gao
  email: gaoyu19920914@gmail.com
date: "`r Sys.Date()`"
package: miaSim
output: 
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: yes
        toc_depth: 2
        number_sections: true
vignette: >
    %\VignetteIndexEntry{miaSim}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    \usepackage[utf8]{inputenc}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(cache = FALSE,
                        fig.width = 9,
                        message = FALSE,
                        warning = FALSE)
```

# Introduction

`miaSim` implements tools for microbiome data simulation based on the
`SummarizedExperiment` [@SE], `microsim` .

Microbiome time series simulation can be obtained by generalized 
Lotka-Volterra model,`simulateGLV`, and  Self-Organized Instability 
(SOI), `simulateSOI`. Hubbell's Neutral model, `simulateHubbell` is used 
to determine the species abundance matrix. The resulting abundance matrix 
from these three simulation models is applied to `SummarizedExperiment`
object or `TreeSummarizedExperiment` object.

`powerlawA` and `randomA` give interaction matrix of species 
generated by normal distribution and uniform distribution, respectively.
These matrices can be used in the simulation model examples.

`simulationTimes` generates lists of time series that can be specified as 
simulation time and time points to keep in simulated time.


# Installation

### Bioc-release

```{r install-bioc,eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

    BiocManager::install("miaSim")
```

# Models for simulating microbiome data sets 

`simulateGLV` is the generalized Lotka-Volterra simulation model fitted to 
time-series estimates microbial population dynamics and relative rates of 
interaction. The model relies on interaction matrix that represents interaction
heterogeneity between species. This interaction matrix can be generated 
with `powerlawA` or `randomA` functions depending on the distribution method. 

`powerlawA` uses normal distribution to create interaction matrix. 

```{r}
library(miaSim)
A_normal <- powerlawA(n.species = 4, alpha = 3)
```

`randomA` uses uniform distribution to create interaction matrix. 

```{r}
A_uniform <- randomA(n.species = 10, d = -0.4, min.strength = -0.8,
                        max.strength = 0.8, connectance = 0.5)
```

The number of species specified in the interaction matrix must be the same 
amount as the species used in the `simulateGLV` and `simulateSOI` models.
`simulateGLV` and `simulateRicker` functions both use the generalized
Lotka-Volterra model.

```{r}
SEobject <- simulateGLV(n.species = 4, A_normal, t.start = 0, 
    t.store = 1000, stochastic = FALSE, norm = FALSE)
ExampleRicker <- simulateRicker(n.species=4, A_normal, tend=100, norm = FALSE)
```

Time series is added to `simulateGLV` with `simulationTimes` function where 
the time points can be kept and extracted from simulation time as a separate 
list.

```{r,eval=FALSE}
Time <- simulationTimes(t.start = 0, t.end = 100, t.step = 0.5, 
                           t.store = 100)
Time$t.index 
```

`simulateHubbell` includes the Hubbell Neutral simulation model which explains 
the diversity and relative abundance of species in ecological communities.
This model is based on the community dynamics; migration, births and deaths.
Loss in society is either replaced by migration or birth.

Similarly, `simulateHubbellRates` also uses the Hubbell Neutral model.Migration 
is determined by the metacommunity and replacement by birth is measure by 
growth rates. The time between events is the determinant of replacement.

```{r}
ExampleHubbell <- simulateHubbell(n.species = 8, M = 10, I = 1000, d = 50,
                                    m = 0.02, tend = 100)

RatesHubbell <- simulateHubbellRates(community_initial = c(0,5,10),
    migration_p = 0.1, metacommunity_p = NULL, k_events = 1, 
    growth_rates = NULL, norm = FALSE, t.end=1000)
```

The Self-Organised Instability (SOI) model can be found in `simulateSOI` and it
generates time series for communities and accelerates stochastic simulation.

```{r}
ExampleSOI <- simulateSOI(n.species = 4, I = 1000, A_normal, k=5, com = NULL,
                                            tend = 150, norm = TRUE)
```

Stochastic logistic model is used in `simulateStochasticLogistic` to determine 
dead and alive counts in community.

```{r}
ExampleLogistic <- simulateStochasticLogistic(n.species = 5)
```

The consumer resource community simulation model `simulateConsumerResource` 
requires the use of the `randomE` function, which returns a matrix containing 
the production rates and consumption rates of each species. The resulting 
matrix is used as a determination of resource consumption efficiency.

```{r}
ExampleConsumerResource <- simulateConsumerResource(n.species = 2,
    n.resources = 4, eff = randomE(n.species = 2, n.resources = 4))
```

The community simulations result in the `SummarizedExperiment` [@SE] 
class object containing the abundance matrix. Other fields, such 
as rowData containing information about the samples, and colData, 
consisting of sample metadata describing the samples, can be added to the 
`SummarizedExperiment` class object.

# Session info

```{r}
sessionInfo()
```
