---
title: "miaSim: Microbiome Data Simulation"
date: "`r Sys.Date()`"
package: miaSim
output: 
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: yes
        toc_depth: 2
        number_sections: true
vignette: >
    %\VignetteIndexEntry{miaSim}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    \usepackage[utf8]{inputenc}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(cache = FALSE,
                        fig.width = 9,
                        message = FALSE,
                        warning = FALSE)
```

# Introduction

`miaSim` implements tools for microbiome data simulation based on the
`SummarizedExperiment` [@SE], `microsim` .

Microbiome time series simulation can be obtained by generalized 
Lotka-Volterra model,`simulateGLV`, and  Self-Organized Instability 
(SOI), `simulateSOI`. Hubbell's Neutral model, `simulateHubbell` is used 
to determine the species abundance matrix. Function `powerlawA` gives
interaction matrix of species which can be used in the simulation model
examples. The resulting abundance matrix from these three simulation
models is applied to `SummarizedExperiment` object or 
`TreeSummarizedExperiment` object.

# Installation

### Bioc-release

```{r install-bioc,eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

    BiocManager::install("miaSim")
```

### The development version of the package 
[Github](https://github.com/microbiome/miaSim). 

# Models for simulating microbiome data sets 

The generalized Lotka-Volterra simulation model fitted to time-series estimates
microbial population dynamics and relative rates of interaction. The simulation
function is stored in `simulateGLV`

`simulateHubbell` includes the Hubbell Neutral simulation model which explains 
the diversity and relative abundance of species in ecological communities.

The Self-Organised Instability (SOI) model can be found in `simulateSOI` and it
generates time series for communities and accelerates stochastic simulation.

The simulations result in abundance matrix which can be used to form 
`SummarizedExperiment` [@SE] object.

# Session info

```{r}
sessionInfo()
```
