---
title: "miaSim: Microbiome Data Simulation"
date: "`r Sys.Date()`"
package: 
    miaSim
output: 
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: yes
        toc_depth: 2
        number_sections: true
vignette: >
    %\VignetteIndexEntry{miaSim}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(cache = FALSE,
                        fig.width = 9,
                        message = FALSE,
                        warning = FALSE)
```

`miaSim` implements tools for microbiome data simulation based on the
`SummarizedExperiment` [@SE], `microsim` .     

# Installation

To install `miaSim`, install `devtools` first, if it is not installed.
Afterwards use the `install_github` function from `devtools`.

```{r, eval=FALSE}
library(devtools)
devtools::install_github("microbiome/miaSim")
```

# Load *miaSim*

```{r load-packages, message=FALSE, warning=FALSE}
library("miaSim")
```

# Models for simulating microbiome data sets 

The generalized Lotka-Volterra simulation model fitted to time-series estimates
microbial population dynamics and relative rates of interaction.

The Hubbell Neutral simulation model explains the diversity and relative 
abundance of species in ecological communities.

The Self-Organised Instability (SOI) model generates time series for communities
and accelerates stochastic simulation.

The simulations result in abundance matrix which can be used to form 
SummarizedExperiment object.
