---
title: "caseStudy2-NutrientThreshold-TreeSE"
author: "The miaSim package authors."
date: "`r Sys.Date()`"
package: miaSim
output: 
  BiocStyle::html_document:
    fig_height: 7
    fig_width: 10
    toc: yes
    toc_depth: 2
    number_sections: true
    dev: png
    type: cairo
vignette: >
  %\VignetteIndexEntry{caseStudy2-NutrientThreshold}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  fig.width = 9,
  message = FALSE,
  eval=TRUE,
  warning = FALSE)
knitr::opts_knit$set(global.par = TRUE)
# knitr::opts_chunk$set(dev = "cairo_pdf")
```

# Case study with miaSim: Nutrient concentration threshold

This document provides the detailed code required to replicate case
study 2 discussed in Gao _et al._ (2023). Methods in Ecology and
Evolution. DOI:
[10.1111/2041-210X.14129](https://doi.org/10.1111/2041-210X.14129)

For general instructions and examples on using the miaSim package
tools, see the
[vignette](https://microbiome.github.io/miaSim/articles/miaSim.html). `miaSim`
implements tools for microbiome data simulation based on different
ecological modeling assumptions. These can be used to simulate species
abundance matrices, including time series. For a detailed function
documentation, see the [function reference
page](https://microbiome.github.io/miaSim/reference/index.html)


Reference: [Available energy fluxes drive a transition in the
diversity, stability, and functional structure of microbial
communities](https://doi.org/10.1371/journal.pcbi.1006793).

The aim of this case study is to design and demonstrate the existence
of nutrient concentration threshold which limits the beta-diversity of
communities.

To fulfill this aim, we designed a gradient of environments, as well
as a gradient of communities.

## Load dependencies

```{r deps}
library(ggplot2)
library(vegan)
library(reshape2)
library(mia)
library(miaSim)
library(philentropy)
library(cluster)

# Load auxiliary functions from:
# https://github.com/microbiome/miaSim/blob/main/vignettes/funcs.R
source("funcs.R") 
```

This batch of simulations is time-consuming. To reduce the calculation
burden, we have decreased the numbers of environments, resources, and
communities from the original 10 to 5, and made other minor
modifications.

## Set random seed and initial shared parameters

```{r pars}
set.seed(42)
n_species <- 5
n_resources <- 5
E <- randomE(n_species, n_resources, mean_consumption = 1, mean_production = 3)
growth_rates <- runif(n_species)
monod_constant <- matrix(rbeta(n_species*n_resources, 10,10),
                               nrow=n_species,
			       ncol=n_resources)
t_store <- 50
n.instances <- 1 # no stochastic process: no need to repeat
```


## Initialization 

Generate communities.

```{r gen}
n.community <- 5 # you can also try 20 or even 50.
density.community <- 0.8
set.seed(42)
# Check the Rmd source file for gradient.df.generator function source code:
# https://github.com/microbiome/miaSim/blob/main/vignettes/caseStudy2-NutrientThreshold.Rmd
community.initial.df <- gradient.df.generator(n_row = n.community,
                                              n_col = n_species,
                                              density_row = density.community,
					      max_gradient = 0.7,
					      error_interval = 0.1)
dist.community.initial.df <- vegdist(community.initial.df, method = "bray")

# Create TreeSummarizedExperiment object for the initial community
community.initial.tse <- TreeSummarizedExperiment(assays=SimpleList(abundances=t(as.matrix(community.initial.df))))

# Visualze dissimilarities for the initial community
makeHeatmap(as.matrix(dist.community.initial.df),
            title = "dissimilarity matrix",
            x.label = "community.1",
            y.label = "community.2")
```

Initialize shared parameters.

```{r crm}
crm_params <- list(
    n_species = n_species,
    n_resources = n_resources,
    x0 = NULL,
    E = E,
    resources = rep(1,n_resources),
    monod_constant = monod_constant,
    migration_p = 0,
    stochastic = FALSE,
    t_start = 0,
    t_end = 50,
    t_step = 1,
    t_store = t_store,
    growth_rates = growth_rates,
    norm=FALSE)
```


Generate resource gradients.

```{r grads}
resourceConcentration <- 10^seq(0,4,1) # 1 to 10000
n.medium <- 5
density.medium <- 0.8
n_species <- 5
set.seed(42)

resource.initial.df <- gradient.df.generator(n_row = n.medium,
    n_col = n_resources, density_row = density.medium,
    max_gradient = 0.7, error_interval = 0.1)
```


## Simulations



```{r grads2}
# Pick the initial community
initial.community <- assay(community.initial.tse, "abundances")

crmExample <- simulateConsumerResource(
    n_species = n_species,
    n_resources = n_resources,
    E = E,
    x0 = as.numeric(initial.community[,1]), # First sample from the initial community
    resources = as.numeric(resourceConcentration[3]*resource.initial.df[1,]),
    growth_rates = growth_rates,
    monod_constant = monod_constant,
    stochastic = FALSE,
    t_end = 50,
    t_step = 1,
    t_store = 50,
    norm = FALSE)

#makePlot(crmExample$matrix)
tsem <- meltAssay(crmExample, assay.type = "counts", add_col_data = "time")
p <- ggplot(tsem, aes(x=time, y=counts, color = FeatureID)) +
        geom_line(show.legend = TRUE, lwd=0.5) +
        labs(title="Species abundances by time") +
        theme_linedraw() +
        theme(plot.title = element_text(hjust = 0.5, size = 14))
print(p)


#makePlotRes(crmExample$resources)
df <- as.data.frame(metadata(crmExample)$resources)
df$time <- colData(crmExample)$time
dft <-  melt(df, id="time")
names(dft) <- c("time", "resources", "S.t")
p <- ggplot(dft, aes(x=time, y=S.t, col = resources)) +
        geom_line(show.legend = TRUE, lwd=0.5) +
        labs(title="Quantity of compounds by time") +
        theme_linedraw() +
        theme(plot.title = element_text(hjust = 0.5, size = 14))
```



```{r grads3, eval=FALSE}
# Generate simulations and store the final community in community.simulation
# Here the final relative abundance table is basisComposition_prop.

set.seed(42)
resourceConcentration <- 10^seq(0,4,1) # 1 to 10000
n.medium <- 5

# make use of parallel computing provided by package `foreach`
library(foreach)
library(doParallel)
cl <- makeCluster(max(detectCores()/2, 1))
registerDoParallel(cl)
basisComposition <- foreach(resConc = resourceConcentration, .combine = rbind) %:% 
    foreach(medium = seq_len(n.medium), .combine = rbind, .packages = "miaSim") %dopar% {
        crm_params$resources <- as.numeric(resource.initial.df[medium,]*resConc)
        paramx0 <- as.list(as.data.frame(t(community.initial.df)))
        crm_param_iter <- list(x0 = paramx0)
        print(paste("resConc", resConc, "medium", medium))
        crmMoments <- miaSim::generateSimulations(model = "simulateConsumerResource",
                                           params_list = crm_params,
                                           param_iter = crm_param_iter,
                                           n_instances = n.instances,
                                           t_end = 50)
        # pick community composition at the last time point		  
        as.data.frame(do.call(rbind, lapply(crmMoments, function (x) {assay(x, "counts")[, ncol(x)]})))
    }
stopCluster(cl)
basisComposition_prop <- basisComposition / rowSums(basisComposition)
```




## Make UMAP plots

In this step, plot result is stored in umap_CRM_gradient_plot, and
this is visualized in different facets.

```{r umap, eval=FALSE}
resourceConcentration <- 10^seq(0,4,1) # 1 to 10000
n.medium <- 5
n.community <- 5
concentration <- as.factor(rep(resourceConcentration, each = n.medium*n.community))
medium <- as.factor(rep(seq_len(n.medium), each = n.community ,times = length(resourceConcentration) ))
community <- as.factor(rep(seq_len(n.community), times = length(resourceConcentration)*n.medium))

# Visualize with UMAP 

## Provide the community data as TreeSE object
library(scater)
tse <- TreeSummarizedExperiment(
                assays=SimpleList(abundances=t(as.matrix(basisComposition))),
		colData=DataFrame(Medium=medium,
		                      Concentration=concentration,
				      Community=community
				      )
		    )
## Add UMAP
tse <- runUMAP(tse, name = "UMAP", exprs_values = "abundances")

## Plot UMAP
plotReducedDim(tse, "UMAP", colour_by="Medium", shape_by="Concentration")

# Same for compositional abundance data
library(mia)
## add relative abundances;
tse <- transformSamples(tse, assay.type="abundances", method="relabundance")
tse <- runUMAP(tse, name = "UMAP_compositional", exprs_values = "relabundance")
plotReducedDim(tse, "UMAP_compositional", colour_by="Medium", shape_by="Concentration")
# Finally with communities
umap_CRM_gradient_plot <<- plotReducedDim(tse, "UMAP_compositional",
   colour_by="Medium", shape_by="Community", size_by="Concentration")
```


## Visualization of the results

In this part, different visualization of results demonstrate (in
various facets) the gradual change of communities' beta diversity.
The first figure indicates that the initial community composition is
more important than the combinations of initial available resources.

The first sub-figure in the second figure demonstrates that in an
oligotrophic (less available nutrients) environment, communities won't
change much in a given time, whilst the last two sub-figures resemble
each other, implying that the nutrient is no longer the limiting
factor of the beta-diversity of the community. This pattern is further
displayed in the following "curve plot".

In the third figure, the second and the th community always stays more
similar, despite their initial dissimilarity, indicating that they
might belong to one community type. This can be validated by input 20
or even 50 as n.community in this case study: communities turns into
clusters in each sub-figures.

```{r umapplots, eval=FALSE}
# FIXME: the visual output can be polished later.
print(umap_CRM_gradient_plot)
umap_CRM_gradient_plot + facet_grid(size_by ~ ., labeller = label_both)
umap_CRM_gradient_plot + facet_grid(colour_by ~ size_by, labeller = label_both)
umap_CRM_gradient_plot + facet_grid(shape_by ~ size_by, labeller = label_both)
umap_CRM_gradient_plot + facet_grid(shape_by ~ colour_by, labeller = label_both)
```


