---
title: "caseStudy2-NutrientThreshold-phyloseq"
author: "miaSim package authors"
date: "`r Sys.Date()`"
package: miaSim
output: 
  BiocStyle::html_document:
    fig_height: 7
    fig_width: 10
    toc: yes
    toc_depth: 2
    number_sections: true
vignette: >
  %\VignetteIndexEntry{caseStudy2-NutrientThreshold}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  fig.width = 9,
  message = FALSE,
  warning = FALSE)
```

# Introduction

This document provides the detailed code required to replicate case
study 2 discussed in Gao _et al._ (2023). Methods in Ecology and
Evolution. DOI:
[10.1111/2041-210X.14129](https://doi.org/10.1111/2041-210X.14129)

This replication study has been implemented with phyloseq. The TreeSE
package has been since then upgraded to use the TreeSE data container.

For general instructions and examples on using the miaSim package tools, see the [vignette](https://microbiome.github.io/miaSim/articles/miaSim.html). `miaSim` implements tools for microbiome data simulation based on different ecological modeling assumptions. These can be used to simulate species abundance matrices, including time series. For a detailed function documentation, see the [function reference page](https://microbiome.github.io/miaSim/reference/index.html)


# Case study 2: Nutrient concentration threshold

(Reference: [Available energy fluxes drive a transition in the
diversity, stability, and functional structure of microbial
communities](https://doi.org/10.1371/journal.pcbi.1006793))

The aim of this case study is to design and demonstrate the existence
of nutrient concentration threshold which limits the beta-diversity of
communities.

To fulfill this aim, we designed a gradient of environments, as well
as a gradient of communities.

## Setup


Load dependencies

```{r load2, eval=TRUE}
library(ggplot2)
library(vegan)
library(reshape2)
library(umap)
library(miaSim)
library(philentropy)
library(ape)
library(cluster)
```

```{r}
#' Generate multiple simulations and store them in a list
#'
#' This function is useful when generating simulations with different parameters
#' or simulating multiple instances to evaluate stochasticity therein.
#'
#' @param model Character: name of the model to use, must be one of
#' "simulateConsumerResource", "simulateGLV", "simulateHubbellRates", and
#' "simulateStochasticLogistic"
#' @param params_list List: a list containing all parameters used in the
#' aforementionned called model
#' @param param_iter List of parameters to overwrite in different simulations.
#' If NULL, then simulate different instances using the same set of parameters.
#' (default: \code{param_iter = NULL})
#' @param n_instances Integer: number of instances to generate
#' (default: \code{n_instances = 1})
#' @param t_end Numeric: end time of the simulation. If not identical with t_end
#' in params_list, then it will overwrite t_end in each simulation
#' (default: \code{t_end = 1000})
#' @return A list containing multiple simulating results
#' @examples
#'
#' params <- list(
#'     n_species = 10,
#'     n_resources = 5,
#'     E = randomE(
#'         n_species = 10, n_resources = 5,
#'         mean_consumption = 1, mean_production = 3
#'     ),
#'     x0 = rep(0.001, 10),
#'     resources = rep(1000, 5),
#'     monod_constant = matrix(rbeta(10 * 5, 10, 10), nrow = 10, ncol = 5),
#'     inflow_rate = .5,
#'     outflow_rate = .5,
#'     migration_p = 0,
#'     stochastic = TRUE,
#'     t_start = 0,
#'     t_end = 20,
#'     t_store = 100,
#'     growth_rates = runif(10),
#'     norm = FALSE
#' )
#'
#' # Recommended standard way to generate a set of n simulations (n=2 here) from a given model
#' # simulations <- lapply(seq_len(2), function (i) {do.call(simulateConsumerResource, params)})
#'
#' @export
generateSimulations <- function(model,
    params_list,
    param_iter = NULL,
    n_instances = 1,
    t_end = 1000) {
    simulations <- list()
    if (params_list$t_end != t_end) {
        warning("t_end in the variable list not equals to t_end")
        params_list$t_end <- t_end
    }
    if (is.null(param_iter)) {
        for (i in seq_len(n_instances)) {
            # print(paste(i, "of", n_instances,"instances in one set of params."))
            simulation <- do.call(model, params_list)
            # simulation$matrix[,colnames(simulation$model)!= "time"]
            simulations[[length(simulations) + 1]] <- simulation
        }
        return(simulations)
    } else {
        if (!all(names(param_iter) %in% names(params_list))) {
            stop(
                "not recognized parameter(s) :",
                names(param_iter)[!names(param_iter) %in% names(params_list)]
            )
        }
        if (length(unique(unlist(lapply(param_iter, length)))) > 1) {
            stop("parameters in list param_iter are not of the same length.")
        }

        simulations_full <- list()
        for (i in seq_along(param_iter[[1]])) {
            # print(paste(i, "of", length(param_iter[[1]]), "sets of params."))
            params_list_local <- utils::modifyList(
                params_list,
                lapply(param_iter, "[[", i)
            )
            simulations_local <- generateSimulations(
                model,
                params_list_local,
                param_iter = NULL,
                n_instances = n_instances,
                t_end = t_end
            )
            simulations_full <- append(simulations_full, simulations_local)
        }
        return(simulations_full)
    }
}
```

(This batch of simulations is time-consuming. To reduce the
calculation burden, we decreased the numbers of environments,
resources, and communities from the original 10 to 5, and made other
minor modifications.)

Set random seed

```{r}
set.seed(42)
```

Set initial shared parameters

```{r}
n_species <- 5
n_resources <- 5
E <- randomE(n_species, n_resources)
growth_rates <- runif(n_species)
monod_constant <- matrix(rbeta(n_species*n_resources, 10,10),nrow=n_species, ncol=n_resources)
t_store <- 50
n.instances <- 1 # no stochastic process: no need to repeat

```



Define generating function. This function generates a data frame, where each row is arranged in an increasing dissimilarity to the first row.

```{r}
gradient.df.generator <- function(n_row, n_col, density_row, max_gradient, error_interval){
    list_initial <- list()
    dissimilarity.gradient <- seq(from = 0, to = max_gradient, length.out = n_row)
    for (i in seq_len(n_row)){
        print(i)
        if (i == 1){
            row_temp <- rbeta(n_col, 1, 1/n_col)
            col_to_remove <- sample(x = seq_len(n_col), size = n_col-n_col*density_row)
            row_temp[col_to_remove] <- 0
            list_initial[[i]] <- row_temp
        } else {
            while (length(list_initial) < i) {
                row_temp <- rbeta(n_col, 1, 1/n_col)
                col_to_remove <- sample(x = seq_len(n_col), size = n_col-n_col*density_row)
                row_temp[col_to_remove] <- 0
                diff_temp <- abs(vegdist(rbind(list_initial[[1]], row_temp), method = "bray") - dissimilarity.gradient[i])
                if (diff_temp < error_interval) {
                    list_initial[[i]] <- row_temp
                }
            }
        }
    }
    dataframe_to_return <- as.data.frame(t(matrix(unlist(list_initial), ncol = n_row)))
    return(dataframe_to_return)
}
```


Generate communities

```{r}
n.community <- 5 # you can also try 20 or even 50.
density.community <- 0.8
set.seed(42)
community.initial.df <- gradient.df.generator(n_row = n.community, n_col = n_species, density_row = density.community, max_gradient = 0.7, error_interval = 0.1)
dist.community.initial.df <- vegdist(community.initial.df, method = "bray")
```

Load plotting functions

```{r}
makePlot <- function(out_matrix, title = "abundance of species by time", obj = "species", y.label = "x.t"){
    df <- as.data.frame(out_matrix)
    dft <-  melt(df, id="time")
    names(dft)[2] = obj
    names(dft)[3] = y.label
    lgd = ncol(df)<= 20
    ggplot(dft, aes_string(names(dft)[1], names(dft)[3], col = names(dft)[2])) +
        geom_line(show.legend = lgd, lwd=0.5) +
        ggtitle(title) +
        theme_linedraw() +
        theme(plot.title = element_text(hjust = 0.5, size = 14))
}
makePlotRes <- function(out_matrix, title = "quantity of compounds by time"){
    df <- as.data.frame(out_matrix)
    dft <-  melt(df, id="time")
    names(dft)[2] = "resources"
    names(dft)[3] = "S.t"
    lgd = ncol(df)<= 20
    ggplot(dft, aes(time, S.t, col = resources)) +
        geom_line(show.legend = lgd, lwd=0.5) +
        ggtitle(title) +
        theme_linedraw() +
        theme(plot.title = element_text(hjust = 0.5, size = 14))
}
makeHeatmap <-function(matrix.A,
                       title = "Consumption/production matrix",
                       y.label = 'resources',
                       x.label = 'species',
                       midpoint_color = NULL,
                       lowColor = "red",
                       midColor = "white",
                       highColor = "blue"){
    df <- melt(t(matrix.A))
    if (is.null(midpoint_color)) {
        midpoint_color <- 0
    }
    names(df)<- c("x", "y", "strength")
    df$y <- factor(df$y, levels=rev(unique(sort(df$y))))
    fig <- ggplot(df, aes(x,y,fill=strength)) + geom_tile() + coord_equal() +
        theme(axis.title = element_blank()) +
        scale_fill_gradient2('strength', low = lowColor, mid = midColor, high = highColor, midpoint = midpoint_color)+
        theme_void() + ggtitle(title)

    if (ncol(matrix.A)<=10 & nrow(matrix.A)<=10){
        fig <- fig + geom_text(aes(label = round(strength, 2)))
    } else if (ncol(matrix.A)<=15 & nrow(matrix.A)<=15){
        fig <- fig + geom_text(aes(label = round(strength, 1)))
    } else {
        fig <- fig
    }

    fig <- fig + labs(x = x.label, y = y.label)+
        theme_linedraw() +
        theme(plot.title = element_text(hjust = 0.5, size = 14), axis.text.x = element_text(
            angle = 90))

    if (nrow(matrix.A) >= 20){
        # too many species
        fig <- fig + theme(
            axis.title.y=element_blank(),
            axis.text.y=element_blank(),
            axis.ticks.y=element_blank(),
        )
    }
    if (ncol(matrix.A) >= 20){
        # too many resources
        fig <- fig + theme(
            axis.title.x=element_blank(),
            axis.text.x=element_blank(),
            axis.ticks.x=element_blank()
        )
    }
    fig
}
makeUMAP <- function(matrix, n_neighbors=10, min_dist=0.1, gradient=NULL, gradient_title = 'gradient', group=NULL, group2=NULL){
    custom.config = umap.defaults
    custom.config$n_neighbors = n_neighbors
    custom.config$min_dist = min_dist

    df <- as.data.frame(umap(matrix,config = custom.config)$layout)
    df$gradient <- gradient

    if (is.null(gradient)){
        df$gradient <- 1

    }
    colnames(df) = c('UMAP_2', 'UMAP_1', gradient_title)
    if (is.null(group)){
        ggplot(df, aes_string('UMAP_2', 'UMAP_1', color=gradient_title)) +
            geom_point() +
            scale_color_gradient(low="blue", high="red")
    } else {
        if (is.null(group2)){
            ggplot(df, aes_string('UMAP_2', 'UMAP_1', color=gradient_title)) +
                geom_point(aes(color = group)) + theme_bw()
        } else {
            ggplot(df, aes_string('UMAP_2', 'UMAP_1', color=gradient_title)) +
                geom_point(aes(color = group, shape = group2)) + theme_bw()
        }
    }
}

makeHeatmap(as.matrix(dist.community.initial.df),
            title = "dissimilarity matrix",
            x.label = "community.1",
            y.label = "community.2")
makeUMAP(matrix = community.initial.df,
         n_neighbors = 5,
         group = factor(seq_len(n.community)),
         gradient_title = "community")

```

Initialize shared parameters

```{r}
crm_params <- list(
    n_species = n_species,
    n_resources = n_resources,
    x0 = NULL,
    E = E,
    resources = rep(1,n_resources),
    monod_constant = monod_constant,
    migration_p = 0,
    t_start = 0,
    t_end = 50,
    t_step = 1,
    t_store = t_store,
    growth_rates = growth_rates)
```

Generate resource gradients

```{r}
resourceConcentration <- 10^seq(0,4,1) # 1 to 10000
n.medium <- 5
density.medium <- 0.8
set.seed(42)
resource.initial.df <- gradient.df.generator(n_row = n.medium, n_col = n_resources, density_row = density.medium, max_gradient = 0.7, error_interval = 0.1)
```

## test one run

```{r simulateConsumerResource, eval=FALSE, include=FALSE}
crmExample <- simulateConsumerResource(
    n_species = n_species,
    n_resources = n_resources,
    resources = as.numeric(resourceConcentration[3]*resource.initial.df[1,]),
    t_end = 50,
    t_step = 1,
    t_store = 50)
#makePlot(crmExample$matrix)
#makePlotRes(crmExample$resources)
```

Generate simulations and store the final community in community.simulation. In this step, the final relative abundance table is basisComposition_prop

```{r eval=FALSE, include=FALSE}
'set.seed(42)
library(miaSim)
community.simulation <- list()
counter_i <- 1
for (resConc in resourceConcentration) {
    for (medium in seq_len(n.medium)){
        crm_params$resources <- as.numeric(resource.initial.df[medium,]*resConc)
        paramx0 <- as.list(as.data.frame(t(community.initial.df)))
        crm_param_iter <- list(x0 = paramx0)
        print(paste("resConc", resConc, "medium", medium))
        crmMoments <- generateSimulations(model = "simulateConsumerResource",
                                          params_list = crm_params,
                                          param_iter = crm_param_iter,
                                          n_instances = n.instances,
                                          t_end = 50)
        community.simulation[[counter_i]] <- as.data.frame(do.call(rbind, lapply(crmMoments, getCommunity)))
        counter_i <- counter_i +1
    }
}
basisComposition <- do.call(rbind.data.frame, community.simulation)
rm(counter_i, community.simulation)
basisComposition_prop <- basisComposition / rowSums(basisComposition)'
```

## Visualization

Make UMAP plots.

In this step, plot result is stored in umap_CRM_gradient_plot, and
this is visualized in different facets.

```{r, eval=FALSE}
concentration <- as.factor(rep(resourceConcentration, each = n.medium*n.community))
medium <- as.factor(rep(seq_len(n.medium), each = n.community ,times = length(resourceConcentration) ))
community <- as.factor(rep(seq_len(n.community), times = length(resourceConcentration)*n.medium))

#plot the result in a UMAP space
#makeUMAP(basisComposition, group = medium, group2 = concentration, gradient_title = 'Medium')
#umap_CRM_gradient <- umap(basisComposition_prop)
# umap_CRM_gradient <- umap(basisComposition)
#umap_CRM_coor <- as.data.frame(umap_CRM_gradient$layout)
#colnames(umap_CRM_coor) <- c("UMAP_1", "UMAP_2")
'umap_CRM_coor <- cbind(umap_CRM_coor, concentration, medium, community)
umap_CRM_gradient_plot <- ggplot(umap_CRM_coor,
                                 aes(UMAP_1, UMAP_2,
                                     # alpha = concentration,
                                     color = medium,
                                     shape = community)) +
    geom_point() +
    # scale_shape_manual(values = c(0, 1, 2, 5, 6, 8, 15, 16, 17, 18)) +
    scale_shape_manual(values = seq(0, n.community -1 ,1)) +
    scale_alpha_manual(values = seq(0.25, 1, 0.75/(length(resourceConcentration)-1))) +
    theme_bw()'
```


In this part, different visualization of results demonstrate (in
various facets) the gradual change of communities' beta diversity.
The first figure indicates that the initial community composition is
more important than the combinations of initial available resources.

The first sub-figure in the second figure demonstrates that in an
oligotrophic (less available nutrients) environment, communities won't
change much in a given time, whilst the last two sub-figures resemble
each other, implying that the nutrient is no longer the limiting
factor of the beta-diversity of the community. This pattern is further
displayed in the following "curve plot".

In the third figure, the second and the fourth community always stays
more similar, despite their initial dissimilarity, indicating that
they might belong to one community type. This can be validated by
input 20 or even 50 as n.community in this case study: communities
turns into clusters in each sub-figures.

```{r eval=FALSE, include=FALSE}
print(umap_CRM_gradient_plot)

print(umap_CRM_gradient_plot + facet_grid(concentration ~ ., labeller = label_both))

print(umap_CRM_gradient_plot + facet_grid(medium ~ concentration, labeller = label_both))

print(umap_CRM_gradient_plot + facet_grid(community ~ concentration, labeller = label_both))

print(umap_CRM_gradient_plot + facet_grid(community ~ medium, labeller = label_both))
```



"Saturation curve" of average beta-diversity between communities with
community 1

In this part, we'd like to demonstrate that the average distance from
other communities to community 1 will reach to a threshold of
nutrients, after which the average distance won't increase along with
the total concentration of nutrients.

To do so, we first define a function calculating the mean distance to
the first community.



Construct a function taking umap_CRM_coor as df and return the mean
distance

```{r, eval=FALSE}
average_distance <- function(df, res_conc_type, com_type, method = "euclidean"){
    sub_df <- df[df$concentration == res_conc_type & df$community == com_type,]
    combines <- combn(sub_df$medium, 2)
    distances <- NULL
    for (i in seq_len(ncol(combines))) {
        distances[i] <- dist(sub_df[combines[,i], c(1, 2)])
    }
    # print(distances)
    return(mean(distances))
}
# average_distance(umap_CRM_coor, 1, 2)
```


Constructing dataframe to save results

```{r eval=FALSE, include=FALSE}
distance_saturation_data <- data.frame(concentration = integer(),
                                       community = integer(),
                                       average_distance = numeric())

for (res_conc_type in unique(umap_CRM_coor$concentration)){
    for (com_type in unique(umap_CRM_coor$community)){
        ave_dist <- average_distance(umap_CRM_coor, res_conc_type, com_type)
        distance_saturation_data[nrow(distance_saturation_data)+1,] <-
            c(res_conc_type, com_type, ave_dist)
    }
}
# View(distance_saturation_data)
distance_saturation_data$average_distance <- as.numeric(distance_saturation_data$average_distance)
distance_saturation_data$concentration <- as.factor(distance_saturation_data$concentration)
distance_saturation_data$community <- as.factor(distance_saturation_data$community)
distance_saturation_data_plot <- ggplot(distance_saturation_data,
                                 aes(concentration, average_distance,
                                     color = community,
                                     group = community)) +
    geom_line() + geom_point() +
    scale_shape_manual(values = c(0, 1, 2, 5, 6, 8, 15, 16, 17, 18)) +
    labs(x = "resource concentration", y = "average distance between communities in UMAP") +
    theme_bw()

```

Visualize the distance plot

```{r eval=FALSE, include=FALSE}
print(distance_saturation_data_plot)
# ggsave(paste0("CRMgradient_distance_curve_mod.pdf"), plot = distance_saturation_data_plot , dpi = 300, width = 12, height = 10, units = "cm", scale = 2)
```