### Auxiliary functions

The `simulationTimes` function generates lists of time series that can
be specified as simulation time and time points to keep in simulated
time.

```{r auxiliary}
Time <- simulationTimes(t_start = 0, t_end = 100, t_step = 0.5, 
                           t_store = 100)

Time$t_sys[Time$t_index]
```

`Time$t_sys` contains all time points, and `Time$t_index` stores the
selected index for time points. Therefore,
`Time$t_sys[Time$t_index]` returns the selected time points. Note that
the tail would be trimmed when the interval between start and end
cannot be divided exactly by step length with no remainder.



## Generating multiple simulations at once

```{r demo1}
params <- list(
    n_species = 10,
    n_resources = 5,
    E = randomE(
        n_species = 10, n_resources = 5,
        mean_consumption = 1, mean_production = 3
    ),
    x0 = rep(0.001, 10),
    resources = rep(1000, 5),
    monod_constant = matrix(rbeta(10 * 5, 10, 10), nrow = 10, ncol = 5),
    inflow_rate = .5,
    outflow_rate = .5,
    migration_p = 0,
    stochastic = TRUE,
    t_start = 0,
    t_end = 20,
    t_store = 100,
    growth_rates = rep(2, 10),
    norm = FALSE
)

###  restructure the parameter list
param_iter_new <-replicate(8, list(list(x0 = runif(n = 10, min = 0.01, max = 0.02))))

CRMSimus_single_param_new <- list()
CRMSimus_single_param_new <- lapply(
  param_iter_new,
  function(pars){
    do.call(
      what = simulateConsumerResource,
      args = modifyList(params, val = pars)
    )
  }
)
```



2. demo to iterate combinations of parameters e.g. 9 different x0 and 9 different growth_rates

```{r demo2}
param_iter2_new <- replicate(9, list(list(x0 = runif(n = 10, min = 0.01, max = 0.02),
                                          growth_rates = runif(n=10, min = 0.5, max = 1))))

CRMSimus_double_param_new <- list()
CRMSimus_double_param_new <- lapply(
  param_iter2_new,
  function(pars){
    do.call(
      what = simulateConsumerResource,
      args = modifyList(crm_params, val = pars)
    )
  }
)
```


```{r}
## 3 demo multiple instances
CRMsimus_multiple_instances_new <- lapply(
  seq_len(5),
  function(pars){
    do.call(
      what = simulateConsumerResource,
      args = crm_params
    )
  }
)
```


```{r}
## 4&5 demo to iterate multiple instances
CRMSimus_single_param_multiple_instances_new <- lapply(
  rep(param_iter_new, each = 3), 
  # or rep(param_iter_new, times = 3), depending on how you'd like to organize the data
  function(pars){
    do.call(
      what = simulateConsumerResource,
      args = modifyList(crm_params, val = pars)
    )
  }
)

CRMSimus_single_param_multiple_instances_new <- lapply(
  rep(param_iter2_new, each = 4),
  function(pars){
    do.call(
      what = simulateConsumerResource,
      args = modifyList(crm_params, val = pars)
    )
  }
)
```

It is also possible to incorporate phylogenetic tree information into
`TreeSummarizedExperiment` objects. For further details, see:

```{r}
library(TreeSummarizedExperiment)
help("TreeSummarizedExperiment-constructor", package = TreeSummarizedExperiment)
```


```{r hubbell2}
hubbellmodelRates <- simulateHubbellRates(x0 = c(0,5,10),
    migration_p = 0.1, metacommunity_probability = NULL, k_events = 1, 
    growth_rates = NULL, norm = FALSE, t_end=100)

```

```{r glvmodel}
A_normal <- powerlawA(n_species = 4, alpha = 3)
glvmodel <- simulateGLV(n_species = 4, A = A_normal, t_start = 0, 
    t_store = 100, t_end=100, stochastic = FALSE, norm = FALSE)
```


